---
title: "Data Visualization - Picking the right chart types"
author: "Deepak Varughese, MD"
format: revealjs
editor: visual
---

## What is Data Visualization

Data visualization is the graphical representation of information and data. '

The use of visual elements like graphs to provide an accessible way to see and understand patterns in data.

Additionally, it provides an excellent way to present data to non-technical audiences,

## Exploratory Charts vs Explanatory Charts

```{r}
library(tidyverse)

# 1. Define quadrant label positions
quadrant_data <- data.frame(
  label = c("Idea illustration", "Everyday dataviz", 
            "Idea generation", "Visual discovery"),
  x = c(-0.5, 0.5, -0.5, 0.5),
  y = c(0.5, 0.5, -0.5, -0.5)
)

# 2. Build the Matrix
ggplot() +
  # Central Quadrant Labels
  geom_text(data = quadrant_data, aes(x, y, label = label), 
            size = 5, fontface = "plain", color = "grey40") +
  
  # Vertical Axis Line (stops at +/- 0.9, labels are at +/- 1.05)
  geom_segment(aes(x = 0, y = -0.9, xend = 0, yend = 0.9), 
               arrow = arrow(length = unit(0.15, "cm"), ends = "both"), 
               color = "grey30", size = 0.6) +
  
  # Horizontal Axis Line (stops at +/- 0.85, labels are at +/- 1.15)
  geom_segment(aes(x = -0.85, y = 0, xend = 0.85, yend = 0), 
               arrow = arrow(length = unit(0.15, "cm"), ends = "both"), 
               color = "grey30", size = 0.6) +
  
  # Axis Labels (Positioned outside the line segments)
  annotate("text", x = 0, y = 1.05, label = "DECLARATIVE", fontface = "bold", size = 4.5) +
  annotate("text", x = 0, y = -1.05, label = "EXPLORATORY", fontface = "bold", size = 4.5) +
  annotate("text", x = -1.15, y = 0, label = "CONCEPTUAL", fontface = "bold", size = 4.5) +
  annotate("text", x = 1.15, y = 0, label = "DATA-DRIVEN", fontface = "bold", size = 4.5) +
  
  # Titles and Theme
  labs(
    title = "THE FOUR TYPES",
    subtitle = "From Good Charts by Scott Berinato"
  ) +
  coord_cartesian(xlim = c(-1.5, 1.5), ylim = c(-1.2, 1.2)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(face = "bold", size = 18, hjust = 0, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 11, color = "grey40", hjust = 0, margin = margin(b = 20)),
    plot.margin = margin(30, 30, 30, 30)
  )
```

## The Dataset

-   The Palmer Penguins Dataset

-   Collection of morphological measurements for three penguin species observed in the Palmer Archipelago, Antarctica.

-   It was introduced by Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER) program.

-   It is a high-quality, popular dataset for data science and visualization.

-   Out of Comfort Zone - Truely Exploratory

## Key Variables

-   Type of Species (Categorical)
-   Name of Island (Categorical / Spatial)
-   Sex (Categorical)
-   Bill Length (Continuous Variable)
-   Bill Depth (Continuous Variable)
-   Body Mass Index (Continuous Variable)
-   Year (Time)

## Table of data

```{r}
#| echo: false
#| message: false
#| 
library(palmerpenguins)
library(flextable)
library(tidyverse)
library(officer)
library(skimr)
library(hrbrthemes)

penguins_head <- penguins %>% 
  drop_na()

penguins_head
#penguins_head <- head(penguins, 10) %>% 
#  drop_na()

#ft <- flextable(penguins_head) %>%
#  font(fontname = "sans", part = "all") %>%
#  fontsize(size = 11, part = "all") %>% 
  
  # Align text: Species left, numbers right
#  align(j = 1, align = "left", part = "all") %>%
#  align(j = 2:5, align = "right", part = "all") %>%
  
  # Styling the Header
#  bold(part = "header") %>%
 # bg(bg = "#f8f9fa", part = "header") %>%
  #color(color = "#333333", part = "header") %>%
  
  # Add subtle borders (Editorial Style)
  #border_remove() %>%
  #hline_top(border = fp_border(color = "black", width = 2), part = "header") %#>%
  #hline_bottom(border = fp_border(color = "black", width = 1), part = "header"#) %>%
  #hline_bottom(border = fp_border(color = "black", width = 2), part = "body") %>%
  
  # Formatting numbers
#  colformat_double(digits = 1) %>%
  
  # Finishing touches
 # set_caption("Raw Data: Morphological Summary of Palmer Archipelago Penguins") %>%
  #autofit()

#ft
```

## Outline

-   Visualizing a single categorical variable

-   Visualizing multiple categorical variables

-   Visualizing a single continuous variable

-   Visualizing a continuous variable with a categorical variable

-   Visualizing multiple continuous variables

-   Visualizing changes over time

## Visualizing a single categorical variable

```{r}
penguins_head <- head(penguins, 20) %>% 
  drop_na() %>% 
  select(species)

penguins_head


```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false

penguins_clean <- penguins %>% 
  drop_na()

## Categorical Variables alone

ggplot(data = penguins) +
  aes(
    x = fct_rev(fct_infreq(species)),
    fill = (species == "Adelie")
  ) + 
  geom_bar(alpha = 0.6, width = 0.7) +
  geom_text(
    stat = "count", 
    aes(label = after_stat(paste0(count, " (", scales::percent(after_stat(prop), accuracy = 1), ")")),
        group = 1
        ),
    hjust = 1.2,                   
    fontface = "bold",
    size = 3,
    color = "#2c3e50"
  )+
  scale_y_continuous(
    name = "Number of Penguins",
    limits = c(0, 175),
    breaks = seq(0, 150, by = 25),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_x_discrete(
    name = "",
    labels = c("Adelie" = "Adélie", 
               "Chinstrap" = "Chinstrap", 
               "Gentoo" = "Gentoo")
  ) +
  # Assign specific colors: TRUE = Highlight, FALSE = Standard
  scale_fill_manual(
    values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), 
    guide = "none" 
  ) +
  coord_flip() +
  labs(
    title = "Adélie Penguins Comprise the Largest Population Share",
    subtitle = "Distribution of penguin species across Palmer Archipelago (n=344)",
    caption = "Source: palmerpenguins R package | Analysis by Deepak Varughese",
    alt = "Horizontal bar chart showing Adélie penguins at 44%, Gentoo at 36%, and Chinstrap at 20%."
  ) +
  theme_ipsum()
  
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)
library(gtsummary)
library(hrbrthemes)


penguins_clean <- penguins %>% 
  drop_na()

ggplot(data = penguins) +
  aes(
    x = fct_rev(fct_infreq(species)),
    # We use both fill (for the bar) and color (for the point)
    fill = (species == "Adelie"),
    color = (species == "Adelie")
  ) + 
  # 1. The "Stick" (Just a very thin bar)
  geom_bar(stat = "count", width = 0.05, alpha = 0.8) + 
  
  # 2. The "Candy" (Placed at the count)
  geom_point(stat = "count", size = 5) + 
  
  # 3. The Text (Nudged outside the dot)
  geom_text(
    stat = "count", 
    aes(label = after_stat(paste0(count, " (", scales::percent(count / sum(count), accuracy = 1), ")"))),
    hjust = -0.3, # Nudges text to the right
    fontface = "bold",
    size = 3.5,
    color = "#2c3e50" # Dark color since it's on the white background now
  ) +
  scale_y_continuous(
    name = "Number of Penguins",
    limits = c(0, 200), # Extra room for the labels
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_x_discrete(
    name = "",
    labels = c("Adelie" = "Adélie", 
               "Chinstrap" = "Chinstrap", 
               "Gentoo" = "Gentoo")
  ) +
  # Match both fill and color so the stick and dot are the same color
  scale_fill_manual(values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), guide = "none") +
  scale_color_manual(values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), guide = "none") +
  coord_flip() +
  labs(
    title = "Adélie Penguins Comprise the Largest Population Share",
    subtitle = "Distribution of penguin species across Palmer Archipelago (n=344)",
    caption = "Source: palmerpenguins R package | Analysis by Deepak Varughese"
  ) +
  theme_ipsum()
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false

library(dplyr)
library(ggplot2)

# 1. Prepare summarized data
penguin_summary <- penguins %>%
  count(species) %>%
  mutate(perc = n / sum(n))

# 2. Build the chart
ggplot(penguin_summary, aes(x = 2, y = n, fill = (species == "Adelie"))) +
  geom_bar(stat = "identity", color = "white", size = 1) +
  # Use coord_polar to turn the bar into a circle
  coord_polar(theta = "y", start = 0) +
  # Create the "doughnut" hole
  xlim(0.5, 2.5) +
  # Add labels inside the segments
  geom_text(aes(label = paste0(n, "\n(", scales::percent(perc, accuracy = 1), ")")), 
            position = position_stack(vjust = 0.5), 
            color = "white", 
            fontface = "bold", 
            size = 4) +
  scale_fill_manual(
    values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), 
    guide = "none"
  ) +
  labs(
    title = "Adélie Penguins Comprise the Largest Population Share",
    subtitle = "Doughnut chart of species distribution (n=344)",
    caption = "Source: palmerpenguins R package | Analysis by Deepak Varughese"
  ) +
  theme_ipsum(base_family = "sans") +
  # Clean up the chart by removing axes and gridlines
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)
library(gtsummary)
library(hrbrthemes)
library(treemapify)


penguins_clean <- penguins %>% 
  drop_na()

penguin_counts <- penguins %>%
  count(species) %>%
  mutate(perc = n / sum(n))


ggplot(penguin_counts, aes(area = n, fill = (species == "Adelie"), label = species)) +
  geom_treemap(alpha = 0.8, color = "white", size = 2) +
  geom_treemap_text(
    aes(label = paste0(species, "\n", n, " (", scales::percent(perc, accuracy = 1), ")")),
    colour = "white",
    place = "centre",
    fontface = "bold",
    grow = FALSE # Setting grow = TRUE makes the text fill the box
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), 
    guide = "none"
  ) +
  labs(
    title = "Adélie Penguins Comprise the Largest Population Share",
    subtitle = "Treemap visualization of species distribution (n=344)",
    caption = "Source: palmerpenguins R package | Analysis by Deepak Varughese"
  ) +
  theme_ipsum(base_family = "sans") +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) # Remove axes
```

## 

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)

# 1. Summarize data
penguin_summary <- penguins %>%
  count(species) %>%
  mutate(perc = n / sum(n))

# 2. Build the Pie
ggplot(penguin_summary, aes(x = "", y = n, fill = (species == "Adelie"))) +
  # The bar width must be 1 to fill the circle
  geom_bar(stat = "identity", width = 1, color = "white") +
  
  # Bend the bar into a circle
  coord_polar(theta = "y", start = 0) +
  
  # Add labels (Calculated to sit in the middle of slices)
  geom_text(aes(label = paste0(species, "\n", n, "\n(", scales::percent(perc, accuracy = 1), ")")), 
            position = position_stack(vjust = 0.5), 
            color = "white", 
            fontface = "bold", 
            size = 3.5) +
  
  scale_fill_manual(
    values = c("TRUE" = "#e63946", "FALSE" = "#2c3e50"), 
    guide = "none"
  ) +
  labs(
    title = "Adélie Penguins Comprise the Largest Population Share",
    subtitle = "Pie chart visualization of species distribution (n=344)",
    caption = "Source: palmerpenguins R package | Analysis by Deepak Varughese"
  ) +
  theme_ipsum() +
  # Clean up the "Chart" look to make it a pure "Graphic"
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(ggwaffle) # The ggplot2 extension for waffles
library(hrbrthemes)

# 1. Prepare data: ggwaffle needs a "grid" format
# The waffle_iron function transforms your raw data into X/Y coordinates
waffle_data <- waffle_iron(penguins %>% drop_na(), aes_d(group = species), rows = 12)

# 2. Plot using standard ggplot syntax
library(tidyverse)
library(palmerpenguins)
library(ggwaffle)
library(hrbrthemes)

# 1. Calculate percentages so they sum to 100
penguin_perc <- penguins %>%
  drop_na() %>%
  count(species) %>%
  mutate(percent = round(n / sum(n) * 100))

# 2. Expand the data to exactly 100 rows (1 row per 1%)
# uncount() is a handy tidyverse function that replicates rows based on a value
waffle_perc_data <- penguin_perc %>%
  uncount(percent) %>%
  waffle_iron(aes_d(group = species), rows = 10)

# 3. Plot the 10x10 grid
ggplot(waffle_perc_data, aes(x, y, fill = group)) +
  geom_waffle(color = "white", size = 0.5) +
  coord_equal() +
  scale_fill_manual(
    values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc"),
    name = NULL
  ) +
  labs(
    title = "Penguin Population Distribution",
    subtitle = "Each square represents 1% of the total population (n=333)",
    caption = "Source: palmerpenguins | Analysis by Deepak Varughese"
  ) +
  theme_ipsum(base_family = "sans") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  )
```

## Two or more categorical variables

```{r}
penguins_head <- head(penguins, 20) %>% 
  drop_na() %>% 
  select(species, island)

penguins_head

```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)
library(ggtext) # Required for HTML styling in labels

# Prepare the data
island_summary <- penguins %>%
  drop_na() %>%
  count(island, species)

# Define our colors as variables for easy reuse in the HTML subtitle
color_adelie <- "#e63946"
color_chinstrap <- "#2c3e50"
color_gentoo <- "#a8dadc"

# Create the HTML subtitle string
subtitle_text <- paste0(
  "Distribution of <span style='color:", color_adelie, ";'>**Adélie**</span>, ",
  "<span style='color:", color_chinstrap, ";'>**Chinstrap**</span>, and ",
  "<span style='color:", color_gentoo, ";'>**Gentoo**</span> penguins across islands"
)

# Plot
ggplot(island_summary, aes(x = island, y = n, fill = species)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7) + 
  scale_y_continuous(labels = scales::percent) +
  coord_flip()+
  scale_fill_manual(
    values = c("Adelie" = color_adelie, "Chinstrap" = color_chinstrap, "Gentoo" = color_gentoo),
    guide = "none" # This removes the standard legend
  ) +
  labs(
    title = "Island Species Composition",
    subtitle = subtitle_text,
    x = NULL, # Island names on X are self-explanatory
    y = "Percentage of Population",
    caption = "Source: palmerpenguins | Analysis by Deepak Varughese"
  ) +
  theme_ipsum(base_family = "sans") +
  theme(
    # This line tells ggplot to interpret the subtitle as HTML
    plot.subtitle = element_markdown(size = 12, lineheight = 1.2),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)
library(ggtext) # Required for HTML styling in labels

# Prepare the data
island_summary <- penguins %>%
  drop_na() %>%
  count(island, species)

# Define our colors as variables for easy reuse in the HTML subtitle
color_adelie <- "#e63946"
color_chinstrap <- "#2c3e50"
color_gentoo <- "#a8dadc"

# Create the HTML subtitle string
subtitle_text <- paste0(
  "Distribution of <span style='color:", color_adelie, ";'>**Adélie**</span>, ",
  "<span style='color:", color_chinstrap, ";'>**Chinstrap**</span>, and ",
  "<span style='color:", color_gentoo, ";'>**Gentoo**</span> penguins across islands"
)

# Plot
ggplot(island_summary, aes(x = island, y = n, fill = species)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) + 
  coord_flip()+
  #scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("Adelie" = color_adelie, "Chinstrap" = color_chinstrap, "Gentoo" = color_gentoo),
    guide = "none" # This removes the standard legend
  ) +
  labs(
    title = "Island Species Composition",
    subtitle = subtitle_text,
    x = NULL, # Island names on X are self-explanatory
    y = "Number of Penguins",
    caption = "Source: palmerpenguins | Analysis by Deepak Varughese"
  ) +
  theme_ipsum(base_family = "sans") +
  theme(
    # This line tells ggplot to interpret the subtitle as HTML
    plot.subtitle = element_markdown(size = 12, lineheight = 1.2),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)
library(plotly)

# 1. Setup consistent colors
color_adelie <- "#e63946"
color_chinstrap <- "#2c3e50"
color_gentoo <- "#a8dadc"
island_grays <- c("Biscoe" = "#888888", "Dream" = "#AAAAAA", "Torgersen" = "#CCCCCC")

# 2. Prepare Data
data_clean <- penguins %>% drop_na()

df_islands <- data_clean %>%
  count(island) %>%
  mutate(
    ids = as.character(island),
    labels = as.character(island),
    parents = "",
    values = n,
    color = island_grays[as.character(island)]
  )

df_species <- data_clean %>%
  count(island, species) %>%
  mutate(
    ids = paste0(island, "-", species),
    labels = as.character(species),
    parents = as.character(island),
    values = n,
    color = case_when(
      species == "Adelie" ~ color_adelie,
      species == "Chinstrap" ~ color_chinstrap,
      species == "Gentoo" ~ color_gentoo
    )
  )

sunburst_df <- bind_rows(df_islands, df_species)

# 3. Create Plot with Caption
plot_ly(
  data = sunburst_df,
  ids = ~ids,
  labels = ~labels,
  parents = ~parents,
  values = ~values,
  type = 'sunburst',
  branchvalues = 'total',
  insidetextorientation = 'radial',
  marker = list(colors = ~color)
) %>%
  layout(
    title = list(
      text = "<b>Composition of Species on each Island</b><br><sup>Sunburst Chart</sup>",
      x = 0.5,
      font = list(family = "Arial", size = 18)
    ),
    # The Caption Line
    annotations = list(
      list(
        x = 1, y = -0.1, # Positions text at the bottom right
        text = "Source: PalmarPenguins. Viz: Deepak Varughese",
        showarrow = F,
        xref = 'paper', yref = 'paper',
        xanchor = 'right', yanchor = 'auto',
        font = list(size = 10, color = "gray")
      )
    ),
    margin = list(t = 80, l = 0, r = 0, b = 40) # Extra bottom margin for the caption
  )
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(treemapify)

plot_data <- penguins %>%
  drop_na() %>%
  count(island, species)

# Branding colors
color_palette <- c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")

ggplot(plot_data, aes(area = n, fill = species, subgroup = island)) +
  # 1. Main tiles with thin, subtle species borders
  geom_treemap(color = "white", size = 1) +
  
  # 2. Heavy-duty subgroup borders to define the Island "containers"
  geom_treemap_subgroup_border(color = "white", size = 8) + 
  
  # 3. Floating Island labels with a background "glow" for readability
  geom_treemap_subgroup_text(
    place = "bottomleft", # Moving to the bottom helps separate from species labels
    grow = FALSE, 
    alpha = 0.7, 
    color = "black", 
    fontface = "bold.italic",
    size = 24,
    padding.x = grid::unit(5, "mm"),
    padding.y = grid::unit(5, "mm")
  ) +
  
  # 4. Species labels
  geom_treemap_text(
    aes(label = species),
    colour = "white",
    place = "center",
    fontface = "bold",
    size = 14
  ) +
  scale_fill_manual(values = color_palette, guide = "none") +
  labs(
    title = "Composition of Species on each Island",
    subtitle = "Treemap",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_ipsum() +
  theme(
    plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 20)),
    plot.caption = element_text(size = 10, color = "darkgray", hjust = 1)
  )
```

## Visualizing a single continuous variable

```{r}
penguins_head <- head(penguins, 20) %>% 
  drop_na() %>% 
  select(body_mass_g)

penguins_head

```

## 

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)

# 1. Calculate the Stats & Quartiles
# We need the specific 25th, 50th, and 75th percentiles to define groups
stats <- penguins %>%
  drop_na(body_mass_g) %>%
  summarise(
    mean_val   = mean(body_mass_g),
    median_val = median(body_mass_g),
    sd_val     = sd(body_mass_g),
    q1         = quantile(body_mass_g, 0.25),
    q3         = quantile(body_mass_g, 0.75)
  )

# 2. Prepare Data with Quartile Groups
plot_data <- penguins %>% 
  drop_na(body_mass_g) %>% 
  mutate(quartile = case_when(
    body_mass_g <= stats$q1 ~ "Q1",
    body_mass_g <= stats$median_val ~ "Q2",
    body_mass_g <= stats$q3 ~ "Q3",
    TRUE ~ "Q4"
  ))

# 3. Create the Visualization
ggplot(plot_data, aes(x = body_mass_g, fill = quartile)) +
  # Using a small binwidth or specific bins to ensure colors align with data points
  geom_histogram(color = "white", bins = 30, alpha = 0.8) +
  
  # Color Palette for Quartiles
  scale_fill_manual(values = c("Q1" = "#a8dadc", "Q2" = "#457b9d", 
                               "Q3" = "#1d3557", "Q4" = "#023047")) +
  
  labs(
    title = "Penguin Body Mass: Distribution by Quartiles",
    x = "Body Mass (g)",
    y = "Count",
    fill = "Quartile",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_ipsum() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom")
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)

# 1. Calculate the Stats & Quartiles
# We need the specific 25th, 50th, and 75th percentiles to define groups
stats <- penguins %>%
  drop_na(body_mass_g) %>%
  summarise(
    mean_val   = mean(body_mass_g),
    median_val = median(body_mass_g),
    sd_val     = sd(body_mass_g),
    q1         = quantile(body_mass_g, 0.25),
    q3         = quantile(body_mass_g, 0.75)
  )

# 2. Prepare Data with Quartile Groups
plot_data <- penguins %>% 
  drop_na(body_mass_g) %>% 
  mutate(quartile = case_when(
    body_mass_g <= stats$q1 ~ "Q1",
    body_mass_g <= stats$median_val ~ "Q2",
    body_mass_g <= stats$q3 ~ "Q3",
    TRUE ~ "Q4"
  ))

# 3. Create the Visualization
ggplot(plot_data, aes(x = body_mass_g, fill = quartile)) +
  # Using a small binwidth or specific bins to ensure colors align with data points
  geom_histogram(color = "white", bins = 30, alpha = 0.8) +
  
  # Standard Deviation Lines
  geom_vline(xintercept = c(stats$mean_val - 2*stats$sd_val, 
                            stats$mean_val + 2*stats$sd_val), 
             color = "black", linetype = "dotted", size = 0.8) +
  geom_vline(xintercept = c(stats$mean_val - stats$sd_val, 
                            stats$mean_val + stats$sd_val), 
             color = "black", linetype = "dashed", size = 1) +
  
  # Mean Line
  geom_vline(aes(xintercept = stats$mean_val), 
             color = "black", linetype = "solid", size = 1.2) +
  
  # Color Palette for Quartiles
  scale_fill_manual(values = c("Q1" = "#a8dadc", "Q2" = "#457b9d", 
                               "Q3" = "#1d3557", "Q4" = "#023047")) +
  
  # Annotations
  annotate("text", x = stats$mean_val - 150, y = 40, label = "Mean", 
           color = "black", angle = 90, fontface = "bold") +
  
  labs(
    title = "Penguin Body Mass: Distribution by Quartiles",
    subtitle = "Bars colored by Quartile | Black lines indicate Mean and SD (±1σ, ±2σ)",
    x = "Body Mass (g)",
    y = "Count",
    fill = "Quartile",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom")
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)

# 1. Calculate Stats (Same as before)
stats <- penguins %>%
  drop_na(body_mass_g) %>%
  summarise(
    mean_val   = mean(body_mass_g),
    sd_val     = sd(body_mass_g),
    median_val = median(body_mass_g),
    q1         = quantile(body_mass_g, 0.25),
    q3         = quantile(body_mass_g, 0.75),
    min_val    = min(body_mass_g),
    max_val    = max(body_mass_g)
  )

# 2. Compute Density Data
# We create a density object so we can "shade" specific sections under the curve
d <- density(penguins$body_mass_g, na.rm = TRUE)
density_df <- data.frame(x = d$x, y = d$y) %>%
  mutate(quartile = case_when(
    x <= stats$q1 ~ "Q1",
    x <= stats$median_val ~ "Q2",
    x <= stats$q3 ~ "Q3",
    TRUE ~ "Q4"
  ))

# 3. Create the Visualization
ggplot(density_df, aes(x = x, y = y)) +
  # Fill the areas by quartile
  geom_area(aes(fill = quartile), alpha = 0.8) +
  
  # Standard Deviation Lines
  geom_vline(xintercept = c(stats$mean_val - 2*stats$sd_val, 
                            stats$mean_val + 2*stats$sd_val), 
             color = "black", linetype = "dotted", size = 0.6) +
  geom_vline(xintercept = c(stats$mean_val - stats$sd_val, 
                            stats$mean_val + stats$sd_val), 
             color = "black", linetype = "dashed", size = 0.8) +
  
  # Mean Line
  geom_vline(aes(xintercept = stats$mean_val), 
             color = "black", linetype = "solid", size = 1) +
  
  # Colors
  scale_fill_manual(values = c("Q1" = "#a8dadc", "Q2" = "#457b9d", 
                               "Q3" = "#1d3557", "Q4" = "#023047")) +
  
  # Annotations (Adjusted Y position for density scale)
  annotate("text", x = stats$mean_val - 120, y = 0.0003, label = "Mean", 
           color = "black", angle = 90, fontface = "bold", size = 3) +
  
  labs(
    title = "Penguin Body Mass: Density Distribution by Quartiles",
    subtitle = "Shaded by Quartile | Black lines indicate Mean and SD (±1σ, ±2σ)",
    x = "Body Mass (g)",
    y = "Density",
    fill = "Quartile",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Visualizing a continuous variable and a categorical variable

```{r}
penguins %>% 
  select(body_mass_g, species) %>% 
    drop_na() %>% 
  head(15)
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)

# Create the Box Plot
penguins %>%
  drop_na(body_mass_g, species) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  
  # 1. Add the Boxplot layer
  # outlier.color makes it easy to spot data points that are statistically extreme
  geom_boxplot(
    width = 0.5, 
    alpha = 0.7
  ) +
  
  # 2. Add 'jittered' points (Optional but recommended)
  # This shows the actual distribution of individual penguins behind the summary
  geom_jitter(width = 0.15, alpha = 0.2, color = "black") +
  
  # 3. Aesthetics & Colors
  scale_fill_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  
  labs(
    title = "Comparison of Penguin Body Mass by Species",
    subtitle = "Box Plot",
    x = "",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_ipsum() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank() # Clean up the X-axis grid
  )
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)

# Create the Violin Plot
penguins %>%
  drop_na(body_mass_g, species) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  
  # 1. Add the Violin layer
  # trim = FALSE shows the full range of the data including the tails
  geom_violin(trim = FALSE, alpha = 0.5, color = "black") +
  
  # 2. Add a narrow Boxplot inside
  # This provides the exact median and IQR markers within the density shape
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, alpha = 0.8) +
  
  # 3. Add Mean point (Optional - helpful for skewed data)
  stat_summary(fun = "mean", geom = "point", shape = 23, size = 2, fill = "white") +
  
  # 4. Aesthetics & Colors
  scale_fill_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  
  labs(
    title = "Body Mass Distribution: Violin vs. Box Plot",
    subtitle = "Violin width shows density | Inner box shows IQR | White diamond is the Mean",
    x = "Species",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_ipsum() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank()
  )
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
# install.packages("ggridges")
library(ggridges)
library(hrbrthemes)

# 1. Prepare and Plot
penguins %>%
  drop_na(body_mass_g, species) %>%
  ggplot(aes(x = body_mass_g, y = species, fill = species)) +
  
  # geom_density_ridges creates the overlapping 'mountain' shapes
  # scale = 1.5 allows the ridges to overlap slightly for that 'mountain range' look
  geom_density_ridges(
    alpha = 0.7, 
    scale = 1.5, 
    rel_min_height = 0.01, # Cleans up the 'tails' at the ends
    color = "white",
    quantile_lines = TRUE, # Adds quartile lines inside the ridges
    quantiles = 2          # This specifically draws the Median
  ) +
  
  # 2. Aesthetics
  scale_fill_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  
  labs(
    title = "Ridgeline Distribution of Penguin Body Mass",
    subtitle = "Vertical line indicates the Median | Overlap shows range similarities",
    x = "Body Mass (g)",
    y = "Species",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_ipsum() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(), # Removes horizontal lines for a cleaner look
    axis.text.y = element_text(vjust = 0)  # Aligns species names with the ridges
  )
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(ggdist) # Specialized for distribution visualizations
library(hrbrthemes)

penguins %>%
  drop_na(body_mass_g, species) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  
  # 1. The "Cloud" (Density)
  stat_halfeye(
    adjust = 0.5, 
    width = 0.6, 
    .width = 0, 
    justification = -0.3, 
    point_colour = NA
  ) +
  
  # 2. The "Umbrella" (Boxplot)
  geom_boxplot(
    width = 0.12, 
    outlier.shape = NA, 
    alpha = 0.5
  ) +
  
  # 3. The "Rain" (Raw Points)
  stat_dots(
    side = "left", 
    justification = 1.1, 
    binwidth = 40,
    alpha = 0.5
  ) +
  
  # Aesthetics
  scale_fill_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  labs(
    title = "Raincloud Plot: Body Mass Distribution",
    subtitle = "Combining density, boxplot, and raw data points",
    x = "",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  coord_flip() + # Rainclouds usually look better horizontally
  theme_ipsum() +
  theme(legend.position = "none")
```

## Visualizing 2 continuous variables

```{r}
penguins %>%
  select(flipper_length_mm, body_mass_g) %>% 
  drop_na()
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)

# 1. Simple Scatterplot with Overall Trend
penguins %>%
  drop_na(flipper_length_mm, body_mass_g) %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  
  # Add points with a professional, neutral color
  geom_point(color = "#2c3e50", alpha = 0.4, size = 2) +
  
  # Add a non-linear trend line (LOESS) to see the natural curve of the data
  geom_smooth(color = "#e63946", method = "loess", size = 1.5) +
  
  labs(
    title = "General Relationship: Flipper Length vs. Body Mass",
    subtitle = "Aggregated data across all species (n = 342)",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_ipsum() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)

# 2D Filled Density Plot without Legend
penguins %>%
  drop_na(flipper_length_mm, body_mass_g) %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  
  # The original filled contour look
  geom_density_2d_filled(alpha = 0.85) +
  
  # Using a professional palette (Viridis is standard for scientific clarity)
  scale_fill_viridis_d(option = "plasma") + 
  
  labs(
    title = "Penguin Morphological Density",
    subtitle = "Relationship between flipper length and body mass across the population",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_minimal() +
  theme(
    # This removes the legend completely
    legend.position = "none", 
    
    # Clean up grid lines for a more editorial Reuters/WHO style
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey95"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", margin = margin(b = 15))
  )
```

## 

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)

penguins %>%
  drop_na(flipper_length_mm, body_mass_g) %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  
  # 1. Use a more cohesive color scale
  # color = NA removes the distracting borders entirely
  geom_hex(bins = 28, color = NA) + 
  
  # 2. Viridis 'mako' or 'inferno' provides a smoother gradient
  scale_fill_viridis_c(option = "mako", direction = -1) +
  
  labs(
    title = "Morphological Clusters: Flipper Length vs. Body Mass",
    subtitle = "Concentration of penguin measurements using hexagonal binning",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  # 3. Clean Theme adjustments
  theme_ipsum() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16, color = "#222222"),
    plot.subtitle = element_text(size = 11, color = "#555555", margin = margin(b = 20)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#f0f0f0"), # Faint grid lines
    axis.title = element_text(face = "italic", size = 10)
  )
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)

# 1. Clean data and create the scatterplot
penguins %>%
  drop_na(flipper_length_mm, body_mass_g, species) %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  
  # Add raw data points with some transparency to handle overlap
  geom_point(alpha = 0.6, size = 2.5) +
  
  # Add linear regression lines for each species to show the trend
  geom_smooth(method = "lm", se = FALSE, size = 1.2, alpha = 0.8) +
  
  # Use the species color palette you've established
  scale_color_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  
  # Formatting to WHO/Reuters standards: Clean labels and minimal grid
  labs(
    title = "Correlation Between Flipper Length and Body Mass",
    subtitle = "Positive linear relationship observed across all three penguin species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 10, face = "italic")
  )
```

## 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)
library(hrbrthemes)

# 1. Prepare and clean data
plot_data <- penguins %>%
  drop_na(flipper_length_mm, body_mass_g, bill_length_mm)

# 2. Generate Plot with Legend
ggplot(plot_data, aes(x = flipper_length_mm, y = body_mass_g, size = bill_length_mm)) +
  
  # shape = 21 (filled circle) with white border for better overlap visibility
  geom_point(alpha = 0.5, fill = "#457b9d", shape = 21, color = "white", stroke = 0.3) +
  
  # Scale size with a clear range
  scale_size_continuous(
    range = c(1, 10),
    breaks = c(35, 45, 55), # Explicit breaks for clarity
    labels = c("35 mm", "45 mm", "55 mm"),
    guide = guide_legend(
      title = "Bill Length",
      title.position = "top",
      override.aes = list(fill = "#457b9d", alpha = 0.8), # Make legend bubbles clearer
      direction = "horizontal"
    )
  ) +
  
  labs(
    title = "Penguin Dimensions: Flipper, Mass, and Bill Length",
    subtitle = "Larger bubbles indicate longer bill lengths (mm)",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_ipsum() +
  theme(
    legend.position = "bottom",
    legend.box.background = element_rect(color = "#f0f0f0", fill = NA), # Subtle border around legend
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "italic")
  )
```

## Changes in categorical data over time

```{r}
penguins %>%
  select(year, species) %>%
  drop_na()
```

## 

```{r}
library(tidyverse)
library(palmerpenguins)

# 1. Summarize counts by species and year
yearly_counts <- penguins %>%
  count(year, species) %>%
  drop_na()

# 2. Create the Stacked Area Plot
ggplot(yearly_counts, aes(x = year, y = n, fill = species)) +
  geom_area(alpha = 0.6, color = "white", size = 0.3) +
  
  # Ensure the x-axis only shows the specific years (integers)
  scale_x_continuous(breaks = c(2007, 2008, 2009)) +
  
  # Use your established editorial color palette
  scale_fill_manual(values = c("Adelie" = "#e63946", "Chinstrap" = "#2c3e50", "Gentoo" = "#a8dadc")) +
  
  labs(
    title = "Sampling Trends: Palmer Archipelago (2007-2009)",
    subtitle = "Stacked area showing the volume of penguins recorded each year",
    x = "Observation Year",
    y = "Number of Penguins",
    fill = "Species",
    caption = "Source: PalmarPenguins. Viz: Deepak Varughese"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "italic")
  )
```

## Tools

-   R (ggplot2)

-   D3

-   JavaScript

-   Tableau

-   PowerBi

-   Excel
